---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# algstat

__algstat__ is a collection of tools to help do algebraic statistics in R.  Many (but not all) of the tools make use of back-end connections to software used in various math communities, such as [Macaulay2](http://www.math.uiuc.edu/Macaulay2/), [Bertini](https://bertini.nd.edu), and [LattE](https://www.math.ucdavis.edu/~latte/) with [4ti2](http://www.4ti2.de).

# Exact inference with log-linear models

One of the most well-developed parts of the package allows users to perform (conditional) exact tests for log-linear models.  There are several great references on the math behind this, such as [Diaconis and Sturmfels' original paper](http://projecteuclid.org/euclid.aos/1030563990), the [Lectures on Algebraic Statistics](http://smile.amazon.com/Lectures-Algebraic-Statistics-Oberwolfach-Seminars/dp/3764389044/ref=sr_1_1?ie=UTF8&qid=1430536908&sr=8-1&keywords=lectures+on+algebraic+statistics), and [Markov Bases in Algebraic Statistics](http://smile.amazon.com/Markov-Bases-Algebraic-Statistics-Springer/dp/1461437180/ref=sr_1_fkmr0_1?ie=UTF8&qid=1430536933&sr=8-1-fkmr0&keywords=aoki%2C+hada%2C+and+takemura), so we'll keep that discussion to a minimum.

We'll begin by doing Fisher's exact test on a built-in dataset called politics.

```{r loading}
library(algstat)
data(politics)
politics

loglinear(~ Personality + Party, data = politics)
```

Exact inference in algebraic statistics is done using MCMC to sample from the conditional distribution of the data given its sufficient statistics under the model.  Consequently, the p-values estimated are only determined up to Monte Carlo error.  The standard p-value is given under the column "p.value" in the row labeled P(samp).  Since this is a 2x2 table, the exact p-value is easily calculated with `fisher.test()`:

```{r fisher}
fisher.test(politics)
```



## Installation

### Installing algstat

* From CRAN: `install.packages("algstat")` (this is not up-to-date)

* From Github (dev version): 
```R
# install.packages("devtools")
# install.packages("mpoly")
devtools::install_github("dkahle/algstat")
```

### Installing supporting software

Coming soon!  See the links above for direct information.

