---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# algstat

__algstat__ is a collection of tools to help do algebraic statistics in R.  Many (but not all) of the tools make use of back-end connections to software used in various math communities, such as [Macaulay2](http://www.math.uiuc.edu/Macaulay2/) (for algebraic computations), [Bertini](https://bertini.nd.edu) (for solving systems of polynomial equations), and [LattE](https://www.math.ucdavis.edu/~latte/) with [4ti2](http://www.4ti2.de) (for lattice problems and, in particular, the computation of Markov bases).

This brief intro is currently under construction.






# Exact inference with log-linear models

One of the most well-developed parts of the package allows users to perform (conditional) exact tests for log-linear models.  There are several great references on the math behind this, such as [Diaconis and Sturmfels' original paper](http://projecteuclid.org/euclid.aos/1030563990), the [Lectures on Algebraic Statistics](http://smile.amazon.com/Lectures-Algebraic-Statistics-Oberwolfach-Seminars/dp/3764389044/ref=sr_1_1?ie=UTF8&qid=1430536908&sr=8-1&keywords=lectures+on+algebraic+statistics), and [Markov Bases in Algebraic Statistics](http://smile.amazon.com/Markov-Bases-Algebraic-Statistics-Springer/dp/1461437180/ref=sr_1_fkmr0_1?ie=UTF8&qid=1430536933&sr=8-1-fkmr0&keywords=aoki%2C+hada%2C+and+takemura), so we'll keep the technical discussion to a minimum.

We'll begin by doing Fisher's exact test on a built-in dataset called politics.

```{r fisher}
library(algstat)
data(politics)
politics

fisher.test(politics)
```

Since the independence model is log-linear, this exact same procedure can be done with __algstat__. The go-to function here is `loglinear()` (formerly `hierarchical()`):

```{r loglinearIntro}
loglinear(~ Personality + Party, data = politics)
```

Exact inference in algebraic statistics is done using MCMC to sample from the conditional distribution of the data given its sufficient statistics under the model.  Consequently, the p-values estimated are only determined up to Monte Carlo error.  The standard p-value is given under the column `p.value` in the row labeled `P(samp)`.  The analogous asymptotic test can be done in either of two ways.  

The first way uses the `loglin()` function from the __stats__ package.  It outputs the likelihood ratio statistic (`Likelihood G^2` in the output above) and Pearson's chi-squared statistic (`Pearson X^2` above), and you have to calculate the p-value yourself.

```{r loglin}
(loglinMod <- stats::loglin(politics, list(1, 2)))
pchisq(loglinMod$pearson, df = 1, lower.tail = FALSE)
```

The second way is `loglm()` in the __MASS__ package, which is a nice wrapper of `loglin()` (in fact, __algstat__'s `loglinear()` function uses the IPF implementation from `loglin()`, although it doesn't need to).  It's syntax looks identical to `loglinear()`'s above, but `loglinear()` can actually take a wide range of model specification syntaxes.  Here's `loglm()`:

```{r loglm}
MASS::loglm(~ Personality + Party, data = politics)
```












# Installation

## Installing algstat

* From CRAN: `install.packages("algstat")` (this is not up-to-date)

* From Github (dev version): 
```R
# install.packages("devtools")
# install.packages("mpoly")
devtools::install_github("dkahle/algstat")
```

## Installing supporting software

Coming soon!  See the links above for direct information.

