---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# algstat

__algstat__ is a collection of tools to help do algebraic statistics in R.  Many (but not all) of the tools make use of back-end connections to software used in various math communities, such as [Macaulay2](http://www.math.uiuc.edu/Macaulay2/) (for algebraic computations), [Bertini](https://bertini.nd.edu) (for solving systems of polynomial equations), and [LattE](https://www.math.ucdavis.edu/~latte/) with [4ti2](http://www.4ti2.de) (for lattice problems and, in particular, the computation of Markov bases).

This brief intro is currently under construction.






# Exact inference with log-linear models

_Note: this section assumes you have [LattE](https://www.math.ucdavis.edu/~latte/) and [4ti2](http://www.4ti2.de) installed and algstat has registered them._

One of the most well-developed parts of the package allows users to perform (conditional) exact tests for log-linear models.  There are several great references on the math behind this, such as [Diaconis and Sturmfels' original paper](http://projecteuclid.org/euclid.aos/1030563990), the [Lectures on Algebraic Statistics](http://smile.amazon.com/Lectures-Algebraic-Statistics-Oberwolfach-Seminars/dp/3764389044/ref=sr_1_1?ie=UTF8&qid=1430536908&sr=8-1&keywords=lectures+on+algebraic+statistics), and [Markov Bases in Algebraic Statistics](http://smile.amazon.com/Markov-Bases-Algebraic-Statistics-Springer/dp/1461437180/ref=sr_1_fkmr0_1?ie=UTF8&qid=1430536933&sr=8-1-fkmr0&keywords=aoki%2C+hada%2C+and+takemura), so we'll keep the technical discussion to a minimum.

### Fisher's exact test

We'll begin by doing Fisher's exact test on a built-in dataset called politics.

```{r fisher}
library(algstat)
data(politics)
politics

fisher.test(politics)
```

Since the independence model is log-linear, this exact same procedure can be done with __algstat__. The go-to function here is `loglinear()` (formerly `hierarchical()`):

```{r loglinearIntro}
loglinear(~ Personality + Party, data = politics)
```

Exact inference in algebraic statistics is done using MCMC to sample from the conditional distribution of the data given its sufficient statistics under the model.  Consequently, the p-values estimated are only determined up to Monte Carlo error.  The standard p-value is given under the column `p.value` in the row labeled `P(samp)`.  The analogous asymptotic test can be done in either of two ways.  

The first way uses the `loglin()` function from the __stats__ package.  It outputs the likelihood ratio statistic (`Likelihood G^2` in the output above) and Pearson's chi-squared statistic (`Pearson X^2` above), but you have to calculate the p-value yourself.

```{r loglin}
(loglinMod <- stats::loglin(politics, list(1, 2)))
pchisq(loglinMod$pearson, df = 1, lower.tail = FALSE)
```

The second way is the `loglm()` function in the __MASS__ package, which is a nice wrapper of `loglin()` (in fact, __algstat__'s `loglinear()` function uses the IPF implementation from `loglin()`, although it doesn't need to).  It's syntax looks identical to `loglinear()`'s above:

```{r loglm}
MASS::loglm(~ Personality + Party, data = politics)
```


### Fisher's exact test on RxC tables

Doing Fisher's exact test on larger problems is a significantly more complicated problem.  The documentation for `fisher.test()` illustrates how it can be used on RxC tables in general, not just on 2x2 tables.  Here's an example from its documentation drawn from Agresti (2002, p.57):

```{r bigFisher}
Job <- matrix(
  c(1,2,1,0, 3,3,6,1, 10,10,14,9, 6,7,12,11), nrow = 4, ncol = 4,
  dimnames = list(
    income = c("< 15k", "15-25k", "25-40k", "> 40k"),
    satisfaction = c("VeryD", "LittleD", "ModerateS", "VeryS")
  )
)

Job

fisher.test(Job)
```

And the __algstat__ counterpart:
```{r bigFisherLogLinear}
loglinear(~ income + satisfaction, data = Job)
```

Note that the asymptotic test can be performed as well.  The chi-square approximation is actually very good here:
```{r bigFisherLogLm}
MASS::loglm(~ income + satisfaction, data = Job)
```


### Fisher's exact test on multi-way tables

`fisher.test()` does not generalize to multi-way tables and is prone to crashing even in large-celled two-way tables (see `?loglinear` for an example).  Thus, the only way to do exact inference in multi-way tables (in general and in R) is to use an algebraic method.  We'll illustrate this using the drugs dataset from `loglinear()`'s documentation, taken from Agresti (2002, p.322), on which we'll test the no-three-way interaction model:

```{r logLinearMultiWay}
data(drugs)
ftable(drugs)

loglinear(subsets(1:3, 2), data = drugs)
```

Note that here we've used the more concise syntax of facet specification.  Doing the same with `loglm()` looks like this:

```{r logLmMultiWay}
MASS::loglm(~ 1*2 + 2*3 + 1*3, data = drugs)
```




# Installation

## Installing algstat

* From CRAN: `install.packages("algstat")` (this is not up-to-date)

* From Github (dev version): 
```R
# install.packages("devtools")
# install.packages("mpoly")
devtools::install_github("dkahle/algstat")
```

## Installing supporting software

Coming soon!  See the links above for direct information.

